---
title: "RepData_PeerAssessment2"
author: "by Fabio Bianchini"
date: "30/04/2017"
output:
  html_document:
    keep_md: yes
  pdf_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    latex_engine: xelatex
---

```{r Library, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
packages <- c("downloader", "ggplot2","R.utils", "dplyr","tidyverse","stringr","forcats", "knitr")
    for (package in packages) {
        library(package, character.only = TRUE, warn.conflicts = FALSE)
    }
```
##Synopsis
In this analysis it will use the **U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database** to answer this to question:

1. Across the United States, which types of events (as indicated in the ğ™´ğš…ğšƒğšˆğ™¿ğ™´ variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.
The following timelines show the different time spans for each period of unique data collection and processing procedures. 
Select below for detailed decriptions of each data collection type. 
<https://www.ncdc.noaa.gov/stormevents/details.jsp>


##Data processing
#####Loading Raw Data
```{r, Load data, cache=TRUE}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download(url, "Storm_Data.bz2", mode = "wb") #Download dataset from specific URL
bunzip2("Storm_Data.bz2", "Storm_data.csv", remove = FALSE, skip = TRUE) # unzip data file
Storm_Data <- read.csv("Storm_data.csv") #
```

```{r, Storm_Data - data analysis}
dim(Storm_Data) # Original dataset dimension
names(Storm_Data) # Variables name in the orginal dataset
```

#####Process/transform the data into a format suitable for the analysis

```{r, ds1 - Rearrange and rename Dataset columb}
ds1 <- as_tibble(Storm_Data)
# variable must have a unique name in the dataset
names(ds1)[names(ds1)=="STATE__"] <- "STATE_NUM"
names(ds1)[names(ds1)=="LONGITUDE_"] <- "LONGITUDE_E"
names(ds1) <- str_to_lower(names(ds1)) # Force lowercase dataset columb names
names(ds1) <-str_replace(names(ds1), "_+$","") # Remove final underscore from columb names
names(ds1) <- str_replace(names(ds1), "_",".") #
names(ds1)
```

```{r, ds2 - Significant observation dataset}
# Remove the observation with no interest for answer the question for this analysis
ds2 <- ds1[ds1$fatalities > 0 | ds1$injuries > 0 | ds1$cropdmg > 0 | ds1$propdmg > 0,] 
dim(ds2)
```

## Result

#####1. Across the United States, which types of events (as indicated in the ğ™´ğš…ğšƒğšˆğ™¿ğ™´ variable) are most harmful with respect to population health?

The variables of interest, for analazing the impact on population healt are `fatalites` and `injuries`so we create a subset from the original dataset with only the variable of interest.

```{r, Summarazie Summarize fatalities and injuries valure grouped by `evtype`}
# Create a dataset with only the columb/variable of interest to answer this question
ds3 <- select(ds2, fatalities, injuries, evtype)
# Force all `evtypes` to uppercase 
ds3$evtype <- str_to_upper(ds3$evtype)
# replace multiple spaces with single space
ds3$evtype <- gsub(" +", " ", ds3$evtype)
# Summarize fatalities and injuries valure grouped by `evtype`
ds4 <- ds3 %>% group_by(evtype) %>% 
        summarise(tot.fatalities = sum(fatalities), tot.injuries = sum(injuries))

# Dimension for summarized dataset
dim(ds4) # 
```

```{r, fatalities analysis}
# Re-organize the dataset 
fatalities <- arrange(ds4, desc(tot.fatalities))
head(fatalities)
```

**Fatalitis analysis**

For this analysis we will consider only the events with n. of fatalities greater that the mean

```{r, plot total fatalities by Storm}
plot_fatalities <- fatalities[fatalities$tot.fatalities > mean(fatalities$tot.fatalities), ]
nrow(plot_fatalities) # Events with n. of fatalities greater that the mean
ggplot(plot_fatalities, aes(tot.fatalities, fct_reorder(evtype, tot.fatalities))) + geom_point() + labs(title="Total fatalities for storm in US (period 1950 - 2011)", x = "N. fatalities") 
```

The TORNADO event has most harmful impact on public health with n. **`r as.integer(max(fatalities$tot.fatalities))`** total fatalities.

*The first 10th Fatalities events*

```{r, show fatalities table}
library(knitr)
kable(plot_fatalities[1:10,], col.names = c("Type of Events","Total Fatalities", "Total injuries") )
```


**Injuries analysis** 

For this analysis we will consider only the events with n. of injuries greater that the mean

```{r, injuries analysis}
injuries <- arrange(ds4, desc(tot.injuries))
mean(injuries$tot.injuries) # Mean value for injuries
```

```{r}
plot_injuries <- injuries[injuries$tot.injuries > mean(injuries$tot.injuries), ]
nrow(plot_injuries) # Events with n. of injuries greater that the mean
ggplot(plot_injuries, aes(tot.injuries, fct_reorder(evtype, tot.injuries))) + geom_point() + labs(title="Total injuries for storm in US (period 1950 - 2011)", x="N. injuries", y="") 
```

The TORNADO event has most harmful impact on public health with n. **`r as.integer(max(injuries$tot.injuries))`** total injuries.

*The first 10th injuries events*

```{r, show injuries table}
library(knitr)
kable(plot_injuries[1:10,], col.names = c("Type of Events","Total Fatalities", "Total injuries"))
```


#####2. Across the United States, which types of events have the greatest economic consequences?


The variables of interest for analazing the **greatest economic consequences of a Storm event** are `Property damage` and `Crop damage`, so we create a subset from the original dataset with only the variables of interest

```{r}
damage <- select(ds2, evtype, propdmg, propdmgexp, cropdmg, cropdmgexp)

```
Due to the particulary form for storm data damage in the original dataset, we need to convert this variables in a form suitable per the correct analysis and rappresentation.

```{r, Convert the variable}
# Convert cropdmgexp and propdmgexp variables
damage$propdmgexp <- as.character(damage$propdmgexp)
damage$cropdmgexp <- as.character(damage$cropdmgexp)
damage$propdmgexp <- str_to_upper(damage$propdmgexp)
damage$cropdmgexp <- str_to_upper(damage$cropdmgexp)
#
damage$propdmg.value <- 0 # New dataset columb for property damage value
damage[damage$propdmgexp == "K", ]$propdmg.value <- 3
damage[damage$propdmgexp == "M", ]$propdmg.value <- 6
damage[damage$propdmgexp == "B", ]$propdmg.value <- 7
#
damage$cropdmg.value <- 0 # New dataset columb for crop damage value
damage[damage$cropdmgexp == "K", ]$cropdmg.value <- 3
damage[damage$cropdmgexp == "M", ]$cropdmg.value <- 6
damage[damage$cropdmgexp == "B", ]$cropdmg.value <- 7
#
damage$totdmg.value <- 0 # New dataset columb for total damage value 
names(damage)
```

Now valorize the new `total damage value` columb as a `total property damage` and `total crop damage` value summ
```{r, summarize the dataset}
damage$totdmg.value <- damage$propdmg*(10^damage$propdmg.value) + damage$cropdmg*(10^damage$cropdmg.value)
```

```{r}
# Summarize property damage and crop damage valure grouped by `evtype`
ds5 <- damage %>% group_by(evtype) %>% summarise(total = sum(totdmg.value))
plot_damage <- arrange(ds5, desc(total))
# For the plot porpuose we consider only events with total damage value greater that the mean
plot_damage <- plot_damage[plot_damage$total > mean(plot_damage$total), ]
nrow(plot_damage) # Events with total damage amount greater that the mean
```


```{r, plot damage value}
ggplot(plot_damage, aes(total/10^9, fct_reorder(evtype, total))) + geom_point() + labs(title="Total economic damage for storm in US (period 1950 - 2011)", x="Damage (Billions of USD)", y="Storm event type") 
```

The TORNADO event has the greatest economic consequences with **`r as.integer(max(plot_damage$total)/10^9)` Billions of USD** total damage value.


*The first 10th great economic events*

```{r, show damage table}
library(knitr)
plot_damage$total <- as.integer(plot_damage$total/10^9)
kable(plot_damage[1:10,], col.names = c("Type of Events", "Economic damage (Billions of USD)"))
```

### Conclusion
The TORNADO event seems to be the most harmful with respect to population health and have the greatest economic consequences. 
